What caused the crisis of the party system in the 1990s?

The crucial word in the title of this essay is ‘crisis’, a word used to describe ‘a moment of danger or suspense in politics.’ The choice of the word crisis is apt for the framing of discussion around the ‘moment’ of the final decade of the 20th Century. The political system of Italy’s First Republic experienced a number of events, each of which made great demands on the actors within it. I will attempt to show each of the different elements that appear to have had an influence, causing suspense and danger, on the Italian party system in the 1990s. I will show that the 1990s were a ‘watershed decade in Italian political development’ (Bull & Rhodes, 1997, p.1), where structural problems in the party system were exposed due to internal and external conjunctural events.

Analysis of the crisis of the party system requires knowledge of the history of Italy’s development as a state, yet there is not the space here to illustrate fully the political and cultural inheritance of the First Republic. The events of the 1990s had their genesis in history that has long since been analysed. A brief outline of the key aspects of Italy’s socio-political history will be sketched here.

Italy is often described as a nation of dualities. The first of such dualities is the legacy of unification: a divide between the North and South of the country. Another divide is between the sacred and the secular. The strength of Catholicism in Italy is great and its role in society significant. A third cleavage between those ‘pro’ and ‘anti’ ‘the system’ has been heavily influenced by Cold War policies to contain Communism. Additionally, the Fascist regime and its legacy to the First Republic must not be forgotten.  These inherited factors shaped the principal structural political elements of the First Republic, and as such provide a framework for the crisis of the party system in the 1990s.

The principal political parties of the First Republic were sharply distinct from one another in ideology and reflective of the cleavages in society. The Christian Democrats (DC), the principal party of the right, reflected a strong anti-communist political Catholicism. The PCI, the communist party, was the principal party of the left, reflective of anti-system sentiment, its electoral support in the central ‘red belt’ of the country. The Socialist party, PSI, broadly representing the secular centre-left, was pro-system. The political parties of the First Republic were deeply part of Italian society; the DC grew out of the Catholic social organisations permitted under the Fascist regime, and the PCI, was active in resistance to the Fascist regime. Each enjoyed strong support. 

Having outlined the principal parties, the system within which they interacted requires introduction. The constitution of the First Republic was drawn up with cross-partisan participation following World War II. It was adopted in 1948 and formed the basis for political interaction for the next 40 years. Due to the divisions in society, the constitutional provision for proportional representation electoral systems resulted in the formation of coalition governments because no party won a clear majority of seats in parliament. From the first parliament of Italy’s Republic, the DC worked with coalition partners to exclude the Communist party from government. The main requirement for inclusion in a DC coalition was sharing their anti-communist ideology. The exclusion of the Communists – and for the early years of the Republic the Socialists as well – by the DC, through the building of centre-right coalitions led to the creation of a ‘one-party dominant system’ (Koff & Koff, 2000, p32). This collusion to exclude the PCI institutionalised a social division and sowed the seeds for the crisis of the 1990s. 

The post-war democracy of Italy has been characterised by the ‘exceptionally powerful role’ played by the parties (Bufacchi & Burgess, 2001, p4). This style of political mediation is described ‘partitocrazia’. In this model of democratic government, the ‘informal parties, rather than the formal institutions, constitute the main vehicles through which conflict is mediated’ (Ibid.). 

Powerful parties and the need to satisfy coalition partners fomented a culture of ‘‘lottizzazione’, or allotment, whereby all aspects of the public sphere, from senior jobs in state-owned banks to the political control of the RAI, the state broadcasting organisation, are parcelled out among parties’ (Ibid. p95). This allotment of the spoils of the state was in addition to the allocation of cabinet, ministerial and committee positions to coalition government partners. It was not only the public sphere that was to come under the influence of the party system; the Italian state was also heavily involved in large-scale industrial activity, creating partisan tensions in industry. This politicisation of both the public sphere and economic activity by powerful parties became a further component of the crisis of 1990s. 

There is a pejorative connotation of the term ‘partitocrazia’. This arises from the clientelist style of the political system. The true nature of the ‘sottogoverno’ – hidden government – of clientelism was revealed during the 1990s.

With these broad stokes outlining the party system that developed during the First Republic, I will move now to the detail of the crisis that began in the 1980s and continued into the 1990s. The crisis of the party system was manifested in two ways. Firstly, the established parties of the First Republic, the DC, PCI and PSI, disappeared from the political sphere. The era of partiocratic government ended. Secondly, there was the dramatic rise and success of new political parties. But what caused these changes? As mentioned in my introduction, there were internal and external conjunctural events that were of great influence during this period. It is to the principal events and issues of the late 1980s and early 1990s that I now turn, with the aim of showing the causes of the crisis of the party system.

The disappearance of the traditional parties of the First Republic can be attributed to three main causes, one external and two internal. I shall explore these three areas and then draw these threads together for they arose conjuncturally.

The external cause of the crisis in the party system in the 1990s was ideological. The depolarisation of the global political system due to the collapse of the Soviet Union had significant effects on the party system. The ending of the Cold War, the discrediting of Eastern European ‘real existing socialism’, and the fall of the Berlin Wall were each key external catalysts for change in the party system. They ‘undermined the anti-communist cleavage around which much of the political system had been organised’ (Grundle & Parker, p2). The Italian Communist Party underwent profound changes as a result of this external influence. The PCI had for a number of years prior to these events pursued a strategy of ‘euro communism’, and had positioned itself as a democratic non-revolutionary party. Yet the collapse of the Soviet Union was a force for further change. It resulted in a split in the party, its demise and the founding of two new parties: the PDS – the Democratic Party of the Left – and RC – Communist Refoundation.  This was followed by a decline in electoral support for the Left during the early 1990s. The PCI appeared to be ‘greater than the sum of its parts’ because the combined vote for the PDS and the RC in 1992 was nearly 6% less than the PCI vote in 1987 (Parker, p42). For Bufacchi & Burgess there ‘is no doubt that the fall of the Berlin Wall [and] the transformation of the PCI into the PDS … contributed to the eventual undoing of the old ruling parties’ (2001, p88). The ending of the Cold War also had a significant impact on the Christian Democrats. In respect of its assumed role of protecting Italy from Communism, the DC’s reason for existing disappeared with the end of the Soviet Union. 

The first internal cause I will look at for the disappearance of the established parties was financial. It developed because the government was no longer able to support its own weight financially. The permeation of all aspects of society by the partiocratic system depended on the ability of the strongest coalition partner, traditionally the DC, to allocate the spoils of government amongst its partners. It also depended on the spoils having a value and that value being recognised by others. Understanding how the spoils came to loose their value and how the DC became unable to distribute them effectively, that is to say, to their advantage, requires analysis of the ‘sottogoverno’ of clientelist exchange. 

The DC, with its influence in the public and economic spheres through ‘lottizzazione’ and the vast nationalised enterprises, facilitated the growth of a corrupt system of exchange. The ‘spoils system’ became an opportunity for the political parties and individuals to enrich themselves at the expense of the state and, ultimately, the Italian citizen. Party financing was corrupt; a system of financial kickbacks payable proportionally to parties dependent on their coalitional strength, became the way for businesses to win public works tenders. By the 1990s, public deficit spending was causing a financial crisis for the Italian state. With a public administration sinking further into ‘the red’ payments for contracts slowed and the return on investment for businesses making illegal payments in order to win contracts became slower and slower. It was this that caused the value of the spoils to decrease.

An economic crisis loomed large in the early 1990s and was caused by the escalation of public sector deficit spending: that is to say by expenditure that was committed in order to receive a kickback from the businesses providing the public services. The task of dealing with the debt burden and placing Italy on a stable financial footing was instrumental in the changes to the party system. Administrative ‘caretaker’ governments headed by technocrats took steps to resolve the financial crisis. Their domestic financial policies were not popular with Italian citizens, but were effective in reducing the scale of the problem. Most saliently, their ‘brusque reduction in the flow of public funds … weakened links to traditional clienteles’ (Grundle & Parker, 1996, p5).

The second internal cause for the crisis in the party system originated in the judiciary. It combined with the first internal cause, the financial thread outlined above, and had far-reaching effects. Anti-corruption investigations headed by Milan magistrates led to the arrest of businessmen and politicians alike. The peeling away of the corrupt layers of ‘partitocratic’ government in what became known as the ‘Tangentopoli’ – ‘Kickback City’ – investigations, revealed the corrupt nature of the ‘sottogoverno’. By the time the ex-Prime Minister (from August 1983 to April 1987), PSI leader Bettino Craxi, received notice that he was under judicial investigation in December 1992, the number of members of parliament implicated in the corruption scandal was around a hundred. The ‘exposure of systematic corruption…cast discredit on all the established parties, although not in equal measure’ and was for Grundle and Parker the ‘most important reason’ why it was ‘impossible to arrest the decline of the established parties of government’ (Ibid, p4-5).

The DC declined most spectacularly, from ‘dominance to extinction’ (Furlong, 1996, p.60) and its example clearly shows all three causes of the crisis of the party system. The communist threat had gone weakening their ideological unity, an economic crisis had failed to be averted causing the electorate to feel dissatisfaction, and the party had been stained with corruption, purging their ranks from the elite to the local administrators. The party disintegrated during the 1990s. The previously factional yet functional DC was shattered. The Partito Popolar Italiano (PPL) was the major successor to the DC; the remaining fragments of the old party reconstituting themselves into the CDC, Christian Democrat Centre, and the Movimento dei Cristiani Sociali. Splits and re-groupings characterised ex-DC centre-right political interaction in the 1990s.

The crisis of the party system did not lead to the end of the party system in Italy. Several emergent parties grew rapidly during the 1990s and political parties remained the principal opportunity for democratic participation. I shall now make some comments on the rise of these new parties to reinforce the points raised in my discussion of the crisis in the party system.

The new political parties reflect the traditional divisions in society with greater precision than the traditional DC / PCI cleavage. The Northern League, a coalition of regional autonomists, and the National Alliance, formed in part by neo-fascists, illustrates this intensification of divisions. These parties reflect both the geographic and pro/anti ‘the system’ divisions in society, and make little attempt to be ‘catch-all’. The League rapidly moved from the fringes to the centre with a strong result in the 1992 elections. Both parties were partners in the 1994 ‘Freedom Pole’ election-winning coalition.

Additionally, the new political parties were able to capitalise on the simple fact that they were not the old parties; they prompted their ‘mani pulite’ (‘clean hands’): a reference to the corruption that stained the traditional parties of government. However the fragrance of corruption lingers on with investigations into the business conduct of Silvio Berlusconi, leader of Forza Italia!, another new political party.

Forza Italia! burst onto the political scene in January 1994. Founded and financed by the media-magnate Silvio Berlusconi and with the support of his media enterprises, the party quickly captured the political centre-ground in the two months before the general election. It won this election with its ‘Freedom Pole’ centre-right alliance. But Berlusconi did not satisfy the electorate sufficiently with his promised separation of his business and political lives and this ‘partitocratica’ link was a factor in the move from government to opposition at the next election for the party.

A positive outcome of the crisis in the party system is the end of ‘one-party’ dominant government. The 1990s were a ‘watershed decade’ for this reason. Italy was in a transition to a democratic system where the failing of a party to meet the demands of the electorate are punished, and where alternation in the party or parties of government and opposition is the norm.

In summary, three things caused the crisis of the party system: the end of the Cold War, an economic crisis and the investigation of corruption. The crisis exhibited itself in two principal ways, the decline of the established parties, both of government and the permanently excluded communist opposition, and the dramatic rise and success of new political parties and alliances.

There is much discussion amongst academics and political commentators as to whether the changes to the political landscape that occurred during the 1990s amount to the beginnings of a ‘Second Republic’ for Italy. Forza Italia! believe that it does:

‘In the past few years a new Italy has arisen, humble and tenacious, proud and honest, moderate but firm in defence of the principles of liberty, with no past to hide, and that is not afraid to hope and to believe.’
Forza Italia! ‘Carta Dei Valori’


Word count: 2491

Bibliography

Bufacchi, V. & Burgess, S., Italy Since 1989, Macmilian, 2001

Bull, M. & Rhodes, M., ‘Between Crisis and Transformation: Italian Politics in the 1990s’, West European Politics, vol. 20, no1 (1997)

Furlong, P., ‘Political Catholicism and the strange death of the Christian Democrats’ in Grundle, S. & Parker, S. (eds), The New Italian Republic, Routledge, 1996

Grundle, S. & Parker, S. (eds), The New Italian Republic, Routledge, 1996

Koff, S. & Koff, S., Italy: From the First to the Second Republic, Routledge, 2000

Parker, S. ‘Electoral reform and political change un Italy, 1991-1994’ in Grundle, S. & Parker, S. (eds), The New Italian Republic, Routledge, 1996

Other sources

Forza Italia! ‘Carta Dei Valori’,
http://www.forza-italia.it/images/cdv/cartadeivalori.pdf, 3rd March 2005 

Concise Oxford Dictionary, 1932


