Who benefits from foreign direct investment by multinational corporations?

In considering the question posed, ‘Who benefits from foreign direct investment by multinational corporations’, further questions are raised. What is the phenomenon of Foreign Direct Investment (FDI)? Who are the actors involved; the multinational corporations (MNC), host-nations, international financial institutions (IFI) and non-governmental organisations (NGO), among many others, and how do they view FDI? Is FDI a positive force with many benefits or a destructive force with few redeeming characteristics? How is FDI manifest? In what tradition of economic thinking does it have its heritage? How is it criticised?

In order to understand the question posed in the title of this essay further, it will be important to have an understanding of the origins and motives of the principle actor: the multinational corporation. The MNC is of classic liberal economic theory stock. It has grown into its modern form in the capitalist economic system, dominant in the western world, over several hundred years. Aspects of it come from the 'concession companies' of the colonial period. But, as Goddard observes, due to the "tremendous diversity" of MNC, they "resist definition and a simple classification". (Goddard, 2003:441).

Firstly I will consider the economic heritage of MNC. As I have already said, MNC come from the classical liberal economic tradition. The MNC is perhaps the perfect example of entrepreneurial activity by a private actor in a marketplace. Whether the MNC be a producer of goods, the supplier of resources or the provider of a service, the principle of profit as a motivation for activity is common. Profit comes from the adding of value to an activity by an actor and the free recognition of that value by another actor in the act of purchase from the marketplace.

The multinational corporation is "an organisation that has its productive activities in two or more countries" (Goddard, 2003:441); although whether these countries each constitute a unique marketplace or are part of a homogenous global marketplace is the subject of the globalisation debate and out of the scope of this essay. In the classic liberal economic writings of Adam Smith and David Ricardo, each country posses a feature that gives it a natural competitive advantage at one activity over another country. The climate and natural resources are two examples of natural competitive advantages. The distribution of competitive strengths and weaknesses between economic actors located in different nation states is claimed to create a mutually advantageous trade market.

During the colonial period of European history, the imperial nation states of Europe subjugated countries across the world with the motive of providing raw materials for industry and markets for finished goods. Each colonial acquisition had its ‘competitive advantage’ exploited during industrial revolution of the Western world. I will return to this ‘development’ later.

I will now consider what FDI is for MNC in relation to liberal economic theory and the gaining of competitive advantage. FDI can be described as the use of capital to acquire an asset overseas by an actor. For multinational corporations, investments are made for strategic reasons: the gaining of a competitive advantage. For MNC, FDI gives the opportunity to gain a competitive advantage in the marketplace.

However, the MNC is very different from the free actor in a marketplace as written about by Smith. The father of liberal economic thought, Smith envisaged a marketplace where by freedom of choice and through the motive of profit, individual economic actors traded with each other and, by some ‘invisible hand’, inequalities are rectified and the society as a whole benefits.

Is the choice of FDI over inward investment – investment in the domestic marketplace – the result of an opportunity cost calculation? Are MNC, seeking to gain a competitive advantage in the marketplace, choosing to export capital from their domestic market to a foreign market as the cost of gaining the advantage sought is lower in the foreign market place? Are MNC seeking to exploit the competitive advantage of one nation for their benefit, i.e. profit, in the same manner of the colonial nation states of Europe?

In order to answer these questions and to gain a further understanding of ‘who benefits’ from the decisions of MNC, I will consider FDI from a structural International Political Economy (IPE) perspective. The structural perspective is rooted in the writings of Marx and Lenin. Writing against the capitalist economic system of Europe, the writings of Marx are of great importance as they provide a perspective from which to evaluate the liberal economic system of today with clarity (Balaam & Veseth, 2005:22).

Marx expounded an economic system based on collectivism in direct opposition to the liberal individual-centric economic system of capitalism. Lenin, building on the writings of Marx, proposed a view that capitalism was The Highest Stage of Imperialism (Lenin, 1916). The exploitation of one nation states competitive advantage by another during the colonial period I referred to earlier was, for Lenin, being replaced with exploitation by capitalist businesses. For Lenin, economic imperialism had replaced military imperialism, leading to what Stephen Hymer was later to describe as the “development of underdevelopment” (Balaam & Veseth, 2005:384), a theme to which I will return.

In taking a structural viewpoint an initial answer to the title of this essay and to some of the questions posed above is that FDI benefits the MNC, who are seeking to exploit a competitive advantage of a nation state for financial gain. However this is overly simplistic, even as a basic definition, as it fails to take into account the views of other actors involved.

As noted in my introduction, there are many actors involved in FDI. The investment in foreign markets by MNC is not only a push transaction; FDI is often pulled by nation states. Governments can be found competing to attract foreign capital investment. The World Bank and the International Monetary Fund (IMF), among other international financial institutions, encourage nation states to become host to MNC, promoting FDI as a solution to economic and social problems. NGO monitor and report on the experiences of FDI. Some are highly critical of this liberal trade practice, seeing the ‘invisible hand’ as inequitably distributing wealth. Let us consider these other actors now.

There are inequalities in the world. Some have more than others. The gap between those that ‘have’ and those that ‘have not’ is enormous. Those among the most deprived in the world live in nation states that were subjugated by imperialism during the colonial period when the under-development of their economies began – countries of ‘the global south’. Decolonisation and the end of the imperial order hardly changed their economic status. Is the ’invisible hand’ of the modern liberal trade market slow with the equitable redistribution of wealth to the benefit of all in a modern global society? Or rather has the ‘development of under-development’ been perpetuated by the neo-imperial economic practice of FDI?

The World Bank and the United Nations (UN) are examples of international organisations working to reduce the scales of inequalities in the world. In October 2004, the Heads of State for 191 members of the UN pledged 8 Millennium Development Goals (MDG) to be reached by 2015. Each goal was specific, measurable and achievable, promising real reductions in global inequality. The annual World Development Report (WDR) of the World Bank is a guide to the “economic, social and environmental state of the world”1. The 2004 report focused on the challenge of meeting the UN MDG.  Titled ‘Making Services Work for Poor People’, the 2004 WDR “focuses on basic services, particularly health, education, water, and sanitation” (World Bank, 2004:xv) as areas where the global poor are disadvantaged compared to the global ‘well off’.

The World Bank sees the involvement of the state in the delivery of basic services as inappropriate and inefficient. In the tradition of classic liberal thinking, the state has no business in business and the World Bank upholds this view. The neo-liberal monetarist policies of the World Bank and the IMF have encouraged governments of nations in the global south to disengage from the provision of public services, confident that the market will provide the same services, but in a more efficient and directly accountable manner.

I feel that to best answer the question ‘Who benefits from foreign direct investment by multinational corporations’, a brief examination of the policy recommendations of the World Bank in the 2004 WDR, along with the report from War On Want, a non-governmental organisation, is needed. In doing so, I intend to reveal the economic theories behind the pro-FDI policy of reducing the role of government, and provide, from a review of literature, an answer.

The World Bank World Development Report of 2004 promotes fiscal and social policies that they consider ‘pro-poor’. These policies have been specifically designed with meeting the 8 development goals of the UN. They are by no means the only solutions to the MDG challenge, but given that they originate from the world’s most influential financial institution and are supported by the IMF, they can be taken as an authoritative measure of the neo-liberal economic climate in which we live.

War on Want is an example of a non-government organization that monitors and reports on the experiences of foreign direct investment of the peoples in the global south. Their work supports people in their struggled to find their own path of development, by working in partnership with progressive governments and organisations, labour groups and international solidarity movements to find solutions to what they view as the failure of the world economic system to deliver a more equal distribution of wealth.

The basic premise of the WDR from the World Bank is that poor people – defined as those who live on less than $1 US per day – are failed by the involvement of government in the delivery of basic needs: clean water, access to sanitation, education and health care. The report advocates a strategy of replacing public service delivery by the state with private business, privatisation, arguing that the market will provide better quality services more efficiently than the state.

The World Bank gives four examples of how the state fails in meeting the needs of poor people: state spending supports those who are non-poor, spending does not reaching the front-line provider, corruption and failings in front-line provision, and lack of demand (WDR, 2004:3-4). Smith and Ricardo, when emerging the ideas of liberalism, observed these failings. The market, argues the World Bank, would not have these failings, as inefficient providers of services would be squeezed out of marketplace by competition from other providers, to the benefit of both the service user and the service provider.

The privatisation of public services has opened up new markets for private enterprise. Those most keen to operate in these new markets have been MNC who have made massive FDI commitments to expand their operations.

For the World Bank, everyone benefits; poor people receive better services, social and economic problems of nations are addressed, private enterprise profits and the United Nations MDG are achieved bringing about a fairer and just society. The ‘invisible hand’ redistributes wealth to all.

War on Want have criticised the policies of the World Bank, arguing the MNC and FDI do not bring about the fair and just society envisaged. They argue that the privatisation of public services has failed the poor. War on Want identify a “basic tension” common across privatisation experiences: “private companies are called upon to respond first and foremost to considerations of profit maximisation, and this inevitably conflicts with the requirements of public service.” (War on Want, 2004a:5).

In their report Profiting from Poverty (2004), War on Want highlight the failings of MNC in the delivery of services; tariff increases causing household budget substitution, unemployment from lay-offs of public sector workers, weak regulatory frameworks, reduced workers’ rights, lower environmental standards and lack of incentive for private enterprise to supply the poor. Foreign direct investment is shown to stifle domestic entrepreneurial activity that might have taken the place of the state.

Another report from War on Want, Foreign Direct Investment: the case for development (2004), challenges the assumption that FDI is “an engine for pro-poor growth” (War on Want, 2004b:6). Case studies of FDI experiences show that it has not provided “stable, long-term economic growth in many countries. Instead governments have allowed companies to enter countries without regulation. These companies have enjoyed huge tax benefits under regimes of lax environmental and labour standards. There is mounting evidence to suggest that increases in FDI are leading to further pressures to deregulate labour markets.” (Ibid).

To summarise before I draw my conclusion, foreign direct investment is a liberal economic trade practice whereby an economic actor exports capital from their domestic market to another market to acquire an asset. Multinational corporations make FDI in order capitalise – to make profit from – the competitive advantage of operating in many market places. Historically, the export of capital was seen as economic imperialism and the contemporary experiences of the Global South would seem to support the view that modern FDI by MNC is neo-imperialist. Support for private enterprise is to strengthen the freedom of the individual against the restraining and inefficient intervention of the state.

Neo-liberal economic theory holds that all benefit from foreign direct investment by multinational corporations due to the nature of the competitive marketplace. The structural perspective, with its Marxist heritage, shows that the only actor to appreciate any return from FDI is the investor who profits from the economic subjugation of others.

In conclusion, it is difficult to draw a single answer to the question posed: who benefits? Many different actors benefit in many different ways from FDI. If the neo-liberal dogma of the trickle down of wealth were in practice a swift rising tide lifting all boats, then the ‘invisible hand’ could be seen working to redistribute wealth to the benefit of all. However, competitive advantage favours those with existing strength. The scale of inequalities in the world, and the path that has led to them, indicates to me that FDI perpetuates the ‘development of under-development’, leading me to conclude that foreign direct investment does not benefit any actor other than the neo-imperial multinational corporation.

Bibliography

Goddard C., Cronin, J., & Dash, K. International Political Economy Basingstoke: Palgrave Macmillan, 2003.

Balaam, D., & Veseth, M., Introduction to International Political Economy, New Jersey: Pearson Education, 2005.

Lenin, V.I., Imperialism, The Highest Stage of Capitalism http://www.marxists.org/archive/lenin/works/1916/imp-hsc November 2004.

The World Bank, World Development Report 2004, Washington: Oxford University Press & The World Bank.

War On Want, Foreign Direct Investment: the case for development, http://www.waronwant.org/?lid=8247 November 2004

War On Want, Profiting from Poverty,
http://www.waronwant.org/?lid=8247 November 2004.

