Why is it so difficult to generalise about the policy making process in British central government?

I am first drawn to the assumption in the question. Is it really difficult to make generalisations about the policy-making process in British central government? The breadth of literature on the subject of the 'policy-making process' is great and while there is some agreement, on the whole the various analytical and normative models differ from each other significantly. There is agreement on the existence of a 'policy-making process'. Yet what form that process takes, its description and purpose are widely debated. I am in agreement with the assumption: it is difficult to generalise! The question sets the scope of the discussion on the difficulties generalising about the policy-making process: British central government. With this firmly in sight, I will explore the aspects I think will best demonstrate the difficulties encountered in making generalisations about the policy-making process. The aspects that interest me most are the models of the policy-making and their weaknesses, the insights given by different levels of analysis and the diversity of the participants.

I will begin by outlining a number of different models of policy-making, before explaining why I find the weakness of models a cause for the difficulty in making generalisations. 

“Policy-making is: 'the process by which governments translate their political vision into programmes and actions to deliver ‘outcomes’ – desired changes in the real world’ (Modernising Government White Paper, 1999.)” – Policy Hub website.

How do the different theoretical models of policy-making relate to this self-definition by British central government? The conventional model is an institutional approach. It is essentially a description of a value-free mechanism that enables governments, given a mandate by citizens, to realise their goals with the support of a neutral civil service. The conventional model is the closest to the definition of policy-making above. However this model does not account for a number of other aspects that other models include as central points. The structural approach, essentially a Neo-Marxist approach, would place emphasis on the economic interests, the ‘political vision’ of the ruling classes and would view the delivered ‘outcomes’ of the policy-making process as in their own interest.

Two further models introduce more complexity. They develop the definition above by examining the influences and constraints on the policy-making flowing from actors external to central government. The pluralist and corporatist models highlight the power of non-governmental groups to shape policy. These two models are not only concerned with the power dispersal within a society, they each emphasise a different type of interaction between these groups and the government. Pluralism assigns the government a neutral role of arbitration, with non-governmental groups negotiating and collaborating on policy definition. Corporatism describes the interaction between government and non-governmental groups as an ‘alliance’ where the latter ‘are given a central role in the policy-making process in exchange for exerting pressure upon their members to conform with government decisions’ (Jones, 2004, p.598). Neither the pluralist model nor the corporatist model would fully support the definition of the policy-making process from the White Paper quoted above.

Grant (2000) has criticised both of these approaches. For him pluralism is lacking as a theoretical model as it only examines one level of power in society, the lower level. It fails to examine the upper level, the core and unquestioned assumptions about our society. Corporatist theory and tripartite practice has lost utility for Grant since the ‘abandonment of neo-Keynesian politics’ by British government (p.52).

The party government and Whitehall models of policy-making are variations on the conventional model. These two models support in part the definition from the White Paper. The party government model stresses political parties as the ‘major channel for policy formulation’ (Ibid.). Yet if political parties are the ‘major channel’, then who are the ‘minor channels’? The Whitehall model places emphasis on the Civil Service and their role in translating political vision into programmes. This model contends that due to their permanence, the Civil Service develops particular ‘departmental views’ of their areas, and that initiatives are substantially theirs and not those of their political masters.

The analyses of policy-making as ‘decision-making’ or ‘project management’ highlights the ways decision-makers make decisions. Rational decision making and incrementalism are examples of analytical approaches which have this ‘process’ emphasis: That is to say the ‘process by which governments translate their political vision into programmes and actions to deliver ‘outcomes’’. 

‘In theory there is no difference between theory and practice. In practice there is.’ – Yogi Berra

The numerous models make it difficult to make generalisations about policy-making. Each of the models highlights one aspect of, or perspective on the policy making process. Yet each model, in order to make its generalisations, weakens its own analysis as it cannot take into consideration all of the elements at once. 

The conventional model is in theory the theory and the practice. The structural model criticises the conventional description and looks at the theory of the practice, seeing hidden motivations for the status quo. The two models that look at the dispersal of power do not evaluate the existence of power dispersal; rather the pluralist and corporatist models look at the relationships and interactions between the power bases. 

The Whitehall and party government models are each looking at a different strand of the same rope and in placing emphasis on one aspect of policy-making they miss out the other. The generalisations that these two approaches make, while excluding some aspects, are still helpful in their analysis of policy-making.

The different models make generalising about the process of policy-making in British central government more complex because they can be utilised at different levels of analysis. The micro level of analysis deals with “the role of interests and governments in particular policy decisions”; the macro level examines “broader questions concerning the distribution of power within contemporary society”. The meso-level of analysis looks at linkages between the aspects of the micro and macro levels (Rhodes & Marsh, 1996, p210).

Different levels of analysis of policy-making in British central government will demonstrate different aspects of the subject. A micro level analysis may show how policy was made in a specific example, but it may not show us truths about the whole of the process in British central government. Macro analysis, in attempting to describe policy-making as a whole, may overlook valuable aspects highlighted by specific cases.

The meta analysis by Parsons in ‘Public Policy’ (1999) is an aggregation of the insights offered by the different approaches and different levels of examination of policy-making process. I summarise his meta analysis of policy-making into three broad categories: philosophical, practical and interpersonal. The structural and conventional models are philosophical approaches. They describe and analyse the theory of policy-making. 

Rational decision making and incrementalism are approaches that fall into the practical category. These two place emphasis on the process of policy-making. Essentially stagiest these approaches consider the process of policy-making as a series of discrete components sequenced as a cycle, beginning with an ‘agenda setting’ stage and finishing with a review period which then feeds into the beginning of a new cycle.

The pluralist and corporatist analyses are both examples of the interpersonal analysis category. These approaches consider the actors in the policy-making process and the interactions between them. As mentioned in my introduction, I find the diversity of actors and their various interactions a crucial reason why difficulty is encountered in making generalisations about the policy-making process in British central government. I will outline the actors in the policy-making process and then consider the interactions between them.

The actors in the policy-making process in British central government are numerous. They can, however, be divided into institutional and non-institutional groups. The institutional actors involved in making policy in British central government each perform different roles in the process at different times. Most visible of the actors are the government of the day and its ministers. These actors have legitimate authority from the electoral process to  ‘translate their political vision into programmes’, to make policy. Parliament as the legislature and parliamentary groupings, the Civil Service and the judiciary are other institutional actors in the policy-making process. Different analysis (above) has highlighted the practical roles of these actors in the policy-making process.

I wish to draw special attention to the institutional actors that are found when the scope of ‘British central government’ is taken to include the European dimension. These are, the European Parliament, the European Court of Justice, the European Commission, the European Council and the Council of Ministers. The UK is a member of the European Union and as Bulmer (in Jones eds. 2004, p.755) comments, “the institutional and policy-making apparatus in Brussels has a significant impact upon the way in which the UK is governed.” Membership of the EU has had the effect of ‘supranationalising’ the policy-making process in British central government; policy can come from above the ‘sovereign’ parliament. This supranational dimension further frustrates any generalisations about the policy making process as it widens the net of possibility.

Non-institutional actors have a significant role in the policy-making process. These actors include the general public, quasi-governmental organisations, media, interest groups, cause groups, academic experts and ‘think tanks’. Both the pluralist and corporatist approaches allocate ‘power’ to these groups, power to participate in the policy-making process. It is the interactions between these groups in the formulation of policy that the interpersonal category of analytical approaches are interested in. 

The study of the interpersonal area of the policy-making process has developed a number of helpful terms for describing the interactions between the actors outlined above. The emergence of the terms ‘policy networks’ and ‘policy community’ have had an impact on the ability to make generalisations about policy-making. These terms, used to describe both the groupings of actors and their interactions, are in themselves generalisations based on observed practice. The development of these terms has been explored in great detail (see Rhodes & Marsh, 1996). 

These ‘policy collectivities’ are, in the words of Colebatch, “relatively stable aggregations of people from a wide range of organisations who find themselves thrown together on a continuing basis to address policy problems” (1998, p.23). Different groups find themselves ‘thrown together’ in the policy-making process for different reasons. Groups can also be excluded from the policy-making process. Colebatch discusses the attributes of inclusion in and exclusion from ‘policy collectivities’. He sees three key elements.  Participation, and non-participation, is based on authority, expertise and order (ibid. p.7). Authority frames action, including some and excluding others. The element of authority is seen most in the philosophical models of policy making: the conventional model and the structural model.

Expertise as a requisite for the participation of a group in a policy network is important. Specialist knowledge about a policy area and shared concerns, if not a shared solution, are brought together in issue networks. Expertise groupings clearly illustrate a ‘network’ or ‘community’ aspect to policy making. A government department devoted to an area will probably have established links with professional associations representing practitioners in that area. There may be academic study of this field which creates further pools of specialist knowledge. The general public, as users of the services provided by a government agency, may have grouped themselves together to represent their interests. It is both prudent and practical for these groups to engage with each other; thus creating a policy community, even if ad hoc and without formal recognition.

Order as a basis for participation in policy communities can be understood in two ways. At the micro-level, participation is sought if the inclusion of a group will ensure a consistent, coherent and coordinated application of any policy formulated. Likewise, a group may be excluded if their inclusion will not promote order. At the macro-level, order as a basis for participation can be viewed as being coercive; the participation of groups being sought to ensure the compliance of the groups’ membership with policy formulated.

Also, a group may perceive its-self to be included as a participant in the policy-making process, yet in reality it is excluded. The inclusion of a group in a policy community can be understood in a Machiavellian way; a group can be excluded by inclusion, it’s desires ‘fenced down’ in the ‘process’ of policy-making.

I have used the terms ‘policy network’, ‘issue network’ and ‘policy community’ interchangeably in the preceding outline in order to illustrate the differences in academic discussion. These metaphors help make sense of what is “essentially the same phenomena” (Colebatch, 1998, p.24). Although for Grant (2000) these metaphors are inadequate and leave him grounds for concern as to whether they are the best ‘encapsulation’ of the interrelationships in policy-making.

In summary, this discussion on the different actors and the interactions between them has brought closer an understanding of why it is so difficult to generalise about the policy-making process in British central government. It has demonstrated that there are numerous actors and that their participation is dependent on a number of elements. Additionally, the notion of informal participation in policy-making through ad hoc groups and the idea of ‘exclusion through inclusion’ demonstrate the complexity of the area. Making generalisations about such detailed diversity is difficult.

I have explored some different aspects of policy-making in British central government. In outlining the different models of the policy-making process I have shown that there is indeed a “distinction between analytical and normative theories” as Grant (2000, p.38) observes and that “theory formulation is not value free”. I have shown that different levels of analysis have different usefulness in providing generalisations about the policy-making process.

In conclusion, generalisations are of limited use. It is difficult to make them about the policy-making process in British central government and, firstly, still be describing both the ‘process’ and the ‘artefact’, and, secondly, not create more ambiguity when policy is concerned with creating coherence to paraphrase Colebatch (2000).

Word count 2306
Bibliography

Bulmer, S., in Jones, B., Kavanagh, D., Moran, M. & Norton, P., eds (2004) Politics UK Harlow, Pearson Education 

Colebatch, H. K., (1998) Policy, Buckingham, Open University Press 

Grant, W., (2000) Pressure Groups and British Politics, Basingstoke, Macmillian

Jones, B. in Jones, B., Kavanagh, D., Moran, M. & Norton, P., eds (2004) Politics UK , Harlow, Pearson Education

Parsons, W. (1995) Public Policy, Cheltenham, Edward Elgar

Rhodes, R. A. W. & Marsh, D., ‘The Concept of Policy Networks in British Political Science: Its Development and Utility’ Talking Politics, Spring 1996

Policy Hub Website http://www.policyhub.gov.uk/bpmaking/index.asp, Checked 18th March 2005.
 
