How much power does the chancellor have in German politics?

The question posed by the title of the essay has a number of aspects that require consideration in order to provide an answer. What is power? Can power be measured? What is the role of Chancellor? How is the role given power? I will attempt an examination of these points within the context provided by the question: the German political system. In order to give understanding of the modern political landscape of Germany, some background and context will be given which I will employ in order to show the development of the role of Chancellor in Germanys political system. 

It is to the notion of power that I first turn. Power is a difficult concept to define. The concept could be understood in two ways, power over and power to. It is power in the context of a relationship that concerns students of politics: power over. Power over refers to the ability of a person, A, to compel another person, B, to do something. Obedience and obligation are bound up in the notion of a person exercising power over another person.  Closely related are the ideas of authority and of legitimacy. Where power refers to the ability of a person to do something, authority refers to the right to do something. 

Legitimacy can be understood best as how power and authority are given. In the political arena, power and authority are distributed in a number of different ways, from the use of force to seize and maintain a position of power, to the democratic allocation of power through constitutionally proscribed means. 

How can power be measured? Perhaps an assessment of the power possessed by someone is best understood in relation to its circumstances.  To answer the question ‘How much power…’ an evaluation of the processes, structures, and relationships surrounding the role of the Chancellor is required. I will now begin to consider the power of the German Chancellor within this context.

In Germany political power is distributed among several actors defined in the constitution, the Basic Law, of the nation. The role of Chancellor is one such actor among many. The Chancellor is the leader of the Government, drawn from the lower chamber of the Reichstag. I seek examine the role of the Chancellor in the German state and consider the legitimacy, authority and power of the position in relation to other actors. In order to do so, the origins of the modern German state will be outlined and key points relating to the role of Chancellor will be highlighted.

The modern German political system was brought into being by the nation building of the German people and the Allied Powers after World War II. Following the cessation of hostilities, a critical need was the establishing of a system to fill the political vacuum left by the defeat of the Nazi regime. The Nazi’s had achieved totalitarian dominance of the state of Germany. Their defeat created the necessity for a new political system to be established. Academics perceive the previous political era, that of the Weimar Republic, as having weaknesses that enabled Hitler and the Nation Party of Germany to build on social unrest among the population to assume a supreme position. I shall give a brief account of the Weimar Republic, the Second Reich of the German nation, and the aspects of it that have been criticised.

It is generally agreed among scholars of German politics that the period of the Weimar Republic, from 1919-1933, can be described as being unstable politically. A key part of the Weimar constitution that created this weak state of governance was the problem of a ‘dual-executive’. This problem arose as power and authority to govern were given to the constitutional head of state, the President of Germany, by means of direct election. The President had authority over the appointment of the Chancellor, which weakened the power of the Chancellor. A further constitutional feature that weakened the role of Chancellor was the destructive vote of no confidence. This measure could be used by the Reichstag to dismiss a Chancellor – a motion the Reichstag could agree upon – without the need for agreement on a replacement. The lack of agreement on a candidate-chancellor would delay the business of Reichstag significantly.

The role of Chancellor, having been weakened by these constitutional arrangements, was further diminished by the extreme proportional representation system for the allocation of Bundestag seats among parties after an election. Proportional representation gave no single party a clear majority in the Bundestag, and coalition governments were formed. During the Weimar Republic, the withdrawal of even a junior coalition partner would be sufficient to destabilise government.

It was these constitutional weaknesses that the Basic Law of Germany, as drafted by representatives of the German people and the Allies, sought to remove. The Basic Law defines the role of the Chancellor, the manner of selection of a Chancellor and also constraints upon the role. I shall review these now.

Article 62 of the Basic Law defines the Federal Government as “the Federal Chancellor and the Federal Ministers”. The role of Chancellor is that of leadership of Germany’s Government, formed from the directly elected Bundestag. The position is not directly elected by the people of Germany, rather a candidate is proposed by the President for which a vote is taken amongst the members of the Bundestag (Basic Law Article 63). Roberts describes the presidents role as “simple” because in practice the Bundestag elections demonstrate “which chancellor-candidate can command a majority, since it will have been obvious which coalition is likely to form after the election” (Roberts, 2000:115). The manner of selection of Chancellor, coupled with an indirectly elected president solves the dual-executive problem, but is the problem of weak coalition government also addressed? Is government under the Basic Law still fragile, constrained by coalition? I will return to this question later.

The destructive vote of no confidence has been rewritten in the Basic Law. The President, upon the instruction of the Bundestag, can dismiss the Chancellor. However, this instruction must now come from a constructive no-confidence vote (Basic Law Article 67). A majority of the members of the Bundestag must now not only agree on no confidence in the Chancellor, but also propose and agree a candidate for a replacement. This measure goes someway to strengthen previously fragile coalition government.

I have looked for the legitimacy of the role of the Chancellor and found it in the Basic Law of Germany. The clearest definition of the power of the Chancellor is found in Basic Law Article 65: “The Federal Chancellor shall determine…the general policy guidelines”. This allows the Chancellor to set the policy agenda for the Government and define the constraints within which Federal Ministers run their departments and is known as the Chancellor principle.

The same document gives a clear definition of the authority of the Chancellor. Article 64 allows the Chancellor to propose their own team of Ministers, who need not be members of the Bundestag. Although the Basic Law states that the power to appoint and dismiss Federal Ministers resides with the President, in practice the President acts with a simple formal role, approving the decision of the Chancellor. 

In leading the Government, the Chancellor, as previously noted, is the leader of the party or coalition of parties that commands a majority of the votes in the Bundestag following an election. In determining the general policy direction, the Chancellor is constrained by coalition politics. Goetz notes that political parties do place a restraint on the power of the Chancellor as the “withdrawal of support by one of the political parties that make up the governing coalition inevitably spells the end of the government” (Goetz, 2003:19). The constraint of coalition is mitigated by the discretion of the Chancellor to appoint Federal Ministers of his choosing and Deputies from coalition parties are invited, by convention, to hold cabinet office.

A further example of the power of the office of the Federal Chancellor of Germany can be found in the ability to create, reorganise, merge and set the boundaries of Federal Ministries by means of a single “administrative ordinance” (Goetz, 2003:25). However, it would appear that despite the proscribed authority of the Chancellor to take such actions, these powers are seldom used dramatically. Paterson and Southern (1991:91) see the Chancellors prerogative in ministerial organisation as “considerable power” as it can be used to ensure compliance of coalition parties (see below).

Having considered the formal powers of the Chancellor as ordained in the Basic Law, I shall now consider the informal and less visible powers concentrated in the Chancellor, in particular the aspects that have given rise to the term ‘chancellor democracy’.

The specific boundaries of the Chancellorial role are not defined; there exists only the Chancellor principle (above). By interpretation of Article 65, the Chancellor can wield significant influence in Government. Goetz (2003:31) notes the ways in which the Chancellor is able to do this, citing “strong central co-ordination of government policy, chancellorial control over (and interference with) the activities of ministers, compliant coalition parties, an executive-dominated Parliament and a marginalized opposition”. These factors, Goetz concludes, have made the Chancellorship an “unrivalled centre of power in the political system” (Goetz, 2003:31). 

The authority to determine the policy guidelines (Basic Law 64) of the government strengthens the Chancellor; power in this area in constitutional. The compliance of coalition parties is an informal power that further strengthens the power of the Chancellor by providing the additional seats to ensure a majority of votes in the Bundestag. The normative make-up of government in Germany is coalition government. This can lead to the marginalisation of parliamentary opposition especially in the event of coalitions between the two largest parties. See, for example, the Grand Coalition of the CDU/CSU and the SDP (Paterson & Southern, 1991:84-85).

The ‘unrivalled’ centre can be found in the Chancellery. This is the Federal Office that supports the Chancellor in his role. From this office government policy is co-ordinated and inter-ministerial relations are managed. Relationships with coalition parties are also managed from this office. Considerable resources are placed at the disposal of the Chancellor. The Office has some 500 staff who are organised to reflect the ministries of government. Conradt (2001:1187) describes the Chancellors’ staff, with a right of access to information coffered upon the department, as “among the most influential members of government”.

With the constitutional powers of the Chancellor compound in the Chancellery, the meaning of ‘Chancellor democracy’ can begin to be understood. As previously mentioned the nomination of a Chancellor-candidate by the President of Germany is a simple choice because it will be apparent which personality will be able to lead a majority-holding coalition. Elections for the Bundestag are becoming increasingly dominated by the persona of a parties’ chancellor-candidate. It could be said that, due to the importance of the party-list vote, and personality-centric campaigning, citizens of Germany are directly electing their Chancellor by their vote. This reinforces the legitimacy of the power of the Chancellor.

What further aspects of the political culture of Germany contribute to the power of the Chancellorial role? Goetz (2003:31) notes that the Chancellorships of those who have filled the role since the inaugural post-war Chancellor, Adenauer, have not taken for granted the concentration of the power in the Chancellor and have had to work in a number of areas for support. He notes that the strength of a Chancellor has “depended on their capacity to mobilize constitutional, party political, coalitional, electoral and policy resources.”  Constitutional resources include articles from the Basic Law referring to the authority of the Chancellor and also other constitutional actors such as the Courts. The support of coalition parties, as I have mentioned, can be mobilized through allocation of Ministerial roles to partners. Electoral and policy resources can be mobilized through the Chancellery with its “enormous publicity apparatus” (Paterson & Southern, 1991:96), placing propagandistic communications power at the disposal of the Chancellor.

Support from within their political party has been crucial to the Chancellorial role. With support from his or her party, the power and authority of the Chancellor is reinforced; without support the Chancellor is weakened. Adenauer, the first Chancellor of the new Federal Republic of Germany and with who the character of ‘chancellor democracy’ is associated, was also the Chairman of his party, the CDU. His successor in the role of Chancellor, Erhard, was unable to mobilize party political resources within the CDU while Adenauer was still Chairman and held the Chancellorship for a little over three years to Adenauer’s fourteen. Brandt, the Chancellor of the early years of the 1969-82 SDP-FDP coalition was also Chairman of his party, the SDP.

Having considered where the legitimacy of the Chancellor is found and evaluated some of the constitutional and intuitional resources that must be mobilised in order to support the role, I will summarise my understanding of the power of the Chancellor before concluding. The Chancellorial role, its position of political leadership, is enormously powerful. The Basic Law reinforces this by constitutional measures to ensure that “either a person is or is not Chancellor” (Paterson & Southern, 1991:88) and that that person is able to determine, and be held responsible for, the policy guidelines of Government (Basic Law Article 65).

However, witness the waxing and waning of the influence and effectiveness of the Office of Chancellor since the formulation of the modern Federal Republic of Germany depending on the resources mobilized by the incumbent. The centralisation of power in the Chancellor that characterised the Adenauer era has continued. Recent Chancellors have continued in this mould, with the Kohl tenure (CDU, 1982-1998) described has having seen a “progressive centralization of executive power”. Chancellor Schröder’s (1998-Present) power was reinforced by his election to the leadership of his party, the Social Democratic Party in 1999, his image cemented as the “undisputed leader of his government” (Goetz, 2003:33). This indicates that the power of the Chancellor is not absolute but dependent on a number of factors, not least the personal style of the office holder.

In conclusion, the Chancellorship has constitutional power in its definition. How much power a specific Chancellor has in German politics varies. There is a constitutionally proscribed base for their power, but further power and influence depends on the personality and ability of the office holder.
Bibliography

Conradt, D. (2001), The German Polity, New York, Longman.

Goetz, K. in Padgett, S. Paterson, W. & Smith, G. editors (2003), Developments in German Politics 3, Basingstoke, Macmillan.

Peterson, W. & Southern, S., (1991), Governing Germany, Oxford, Basil Blackwell.

Roberts, G. (2000), German Politics Today, Manchester, Manchester University Press.