Outline what is meant by the two strands of British Conservatism and consider whether they are compatible.

In considering the two strands of British conservatism I am conscious that commentators and critics view this set of principles more as a school of thought than an ‘ism’ or ideology. This lack of a fixed dogma, in contrast to communism for example, makes the job of presenting an outline of the two strands of British conservatism challenging. Yet to this challenge I rise. In order to meet it, I will present my analysis in three parts. First, I shall present some themes in conservative thought. I will then outline each of the strands in conservatism in Britain and how they relate to the themes discussed. Thirdly, I shall look at where the two strands are compatible and examine where some incompatibilities between the two can be found before drawing my conclusions.

Scholars more adept than myself have discussed the main themes in conservatism in great depth. There is not the space here, nor is it my purpose, to explore fully each of the themes. Yet to understand the two strands of British conservatism and assess their compatibility, we need to consider the general themes in conservative thought. It is said that ideologically, the conservative travels light. Of that light ideological baggage, the following five points are foremost. First, the conservative places emphasis on tradition over change. This is not to say that conservatives are descendants of King Canute. Rather, that change should be accepted with prudence. Haywood puts this across succinctly: ‘Tradition, in this sense, reflects the accumulated wisdom of the past’ (Haywood, 1992, p.59).

The ideas of human imperfection and the organic nature of society are closely linked to conservative thought. These two ideas have their roots in the work of Darwin and the theory of evolution and, though this appears contradictory, in the Christian faith. Conservatives are sceptical of the enlightenment project of the 18th century. They reject the idea of social engineering in order to acheive a utopian society. Their view is that the individual is inherently imperfect, innately selfish and responsible of their own ‘bad’ behaviour. These traits of humankind lead to disorder in the conservative view. Yet because they believe that each individual desires security and prosperity, the state, for the conservative, has a coercive role to play in instilling order in a disordered world.

For the conservative, society is organic; a living embodiment of all that has passed, all that is present and all that will come. For Burke, the ‘unprincipled facility of changing the state as often, and as much, and in as many ways as there are floating fancies or fashions’ breaks the linkages in society, leaving man ‘little better than the flies of summer’ (Burke in Rosen & Wolff eds, 1999, p.326). In this view, society is not mechanical and is not to be tinkered with by ‘rational man’ – for the conservative, no such thing exists! Conservatives view the individual as an inseparable part of the organic society. Interactions between individuals and groups give meaning to human existence, providing security. The structures that give this security are to be preserved and respected. Participation in the upkeep of the ‘inheritance of the generations’ is the duty of the individual.

The fourth theme in conservative thought is that of authority. Developing from natural necessity rather than liberal social contract theory, authority is, for the conservative, entrenched in society. Haywood expressed authority as ‘necessary and beneficial as everyone needs the guidance, support and security of knowing ‘where they stand’ and what is expected of them’ (Haywood, 1992, p.65).  With authority a central tenet of conservatism, a paternalistic role is ascribed to the state, with obedience to the state the duty of the citizen. A further element of authority is hierarchy. Conservatives think that hierarchy is necessary to society. What Socialists may describe as social inequality conservatives will describe as hierarchy. For conservatives, social inequality is a fiction and differences in the social positions of individuals are explained as the result of natural due to natural differences in abilities and skills between people. Hierarchies reflect these differences, with the most skilled and able people occupying the highest positions in the hierarchy.

The fifth theme in conservatism is that of property. For conservatives, property is to be owned by an individual, and they reject the collectivisation or the redistribution of property, as advocated by Communists and Socialists respectively. They reject these ideas on the grounds that they would produce a bland and monotonous existence for humankind. Property has benefits for the individual and the state according to the conservative. It gives the individual a stake in society because of his or her desire to protect her or her property and the state a role in protecting the property of individuals by legislation and law enforcement.

I turn now to the two strands of British conservatism. In presenting them, I will outline what is meant by each of the two strands and relate them to the themes discussed above. 

How best to outline what is meant by the two strands of British conservatism? As I mentioned in my introduction, conservatism is more of collection of ideas than an ideology. I shall respect the lack of dogma in the conservative school of thought and keep my terms non-ideological for this simple generalisation: the two strands of British conservatism are the ‘wet’ and the ‘dry’ (Willets, 1992).

The ‘wet’ strand of British conservatism is better known as ‘One Nation’ conservatism. The expression ‘One Nation’ comes from an observation made by the Conservative Prime Minister of the 19th century, Benjamin Disraeli, on the development of two nations, one rich and the other poor. The development of two nations was seen as a threat to the stability of society and the social order. The perception of the potential for social unrest, even revolution, expressed here gives this strand of British conservatism an elitist thread: the status quo is to be upheld and social order preserved by paternalistic intervention in society. In short the assumption is that the interests of the rich nation will serve to protect the poor.

The ‘wets’ hold the principle of noblesse oblige, that is to say that privilege includes responsibility to those unprivileged. While accepting social inequalities due to differing skills and merits of individuals and the hierarchical nature of society, the elements of duty and responsibility of one citizen to another led Disraeli to propose, and win parliamentary approval for, the Social Reform Act of 1867. Disraeli’s Social Reform Act also demonstrates well the conservative notion of, in the words of Burke, ‘change in order to conserve’. In order to strengthen the hierarchy of the state, a stake in making government was given to working class men by the extension of the franchise to them.

Community is a key aspect of ‘wet’ conservatism. In this sense, community means those aspects of our society that are the ties of a group: location, language, tradition and history. Community is central to the identity of the individual for this strand of British conservatism. The shared values of a community sustain a stable social structure – community strengthens the frailty of the human condition and ‘order’ is found within these ties for the paternalist conservative.

The second strand of British conservatism, the ‘dry’ strand, places emphasis on markets rather than communities. As Willets, to whom I owe the wet / dry distinction, writes:

Of course there is more to life, and indeed to politics, than markets. Any conservative understands that, as the previous chapter [titled ‘The Need for Community’], showed. Nevertheless, markets do matter; and they certainly matter to conservatives. It is no accident that almost all the great conservatives thinkers, from David Hume and Edmund Burke through to Michael Oakeshott and Enoch Powell, have been believers in free markets (Willets, 1992, p.79).

Conservative thought was influenced by the free-market writings of the economists Hayek during the 1940s, although conservatives have stressed the need for economic freedom and sound money from before Disraeli. But how is ‘the market’ strand of British conservatism related to the themes already outlined?

Conservatives who emphasise the importance of the market in delivering individual freedom and prosperity see their strategy thus: the human instinct to ‘truck, barter and exchange … leads to spontaneous order’ (Willets, 1992, p.99). Property, for the ‘dry’ conservative, gives the individual a stake in society. The state, for conservatives, should not legislate collectivist or socialised property policies in order to achieve an ordered society; it need only provide regulation and contract enforcement. The state has the authority to do so and uses this authority for the preservation and protection of our organic society.

But ‘free markets’ are not to be laissez-faire for ‘dry’ conservatives, nor are they viewed as being in ‘perfect competition’ or in equilibrium. Markets are viewed as being just as imperfect as the humans that participate in them. But this does not lead the economically focused conservative to support ‘big government’. Why should government intervene? Willets expresses this clearly: ‘People who will ascribe the lowest motives and greatest incompetence to big firms, will then ascribe the highest motives and greatest competence to big government … government decisions are at least as imperfect as those taken in the marketplace … Governments also act on inadequate information’ (1992, p.85). It is for this reason that the Conservative governments of Thatcher and Major in the 1980s and 1990s sought to ‘roll back the state’, privatising nationalised utilities to reduce the governments role in the market.

It is difficult to see how the two strands of British conservatism, ‘One Nation Toryism’ and ‘economic liberalism’, can be compatible or even be reconciled with each other. Are communities restrictive of free markets? Do markets destroy communities? It is to these questions I now turn.

It would appear that these two strands are separated by a deep chasm. While the gap between these two threads is deep, it is not wide; the conservative has a foot on each side. For the conservative, communities need free-markets and free-markets need communities. Communities are the social groups that hold the traditional shared values of a society. They are to be protected by the state for reasons of social cohesion and stability. Free-markets facilitate the accumulation of wealth giving personal freedom and prosperity. This is to be protected by law and regulation rather than intervention. 

Does modern British conservatism uphold both the market and communities? Is the Conservative party of the 21st century both ‘wet’ and ‘dry’?  Willets (1992, p.36), to whom I have turned to again and again for his clarity of expression, sees modern conservatism as both instrumentalist and future-orientated. The instrumentalist approach to government, ‘this is what government can do for you’, is about providing the framework for free-markets rather than being managerial and intervening in the economy. The future-orientated approach to government, ‘this is how we can make things better’, is reflected in welfare state on conservative grounds, i.e. paternalistic, noblesse oblige.

But is conservatism ‘working’ in 21st century Britain? ‘Senior Tories admit the party still has to persuade votes that it could deliver the best of both worlds [communities and free-markets] by reducing taxes while protecting key public services’ The Independent, 15th March 2005.

Word count 1877


Bibliography

Burke, E., in Rosen & Wolff eds, (1999) Political Thought, Oxford, Oxford Readers

Heywood, A., (1992) Political Ideologies, Basingstoke, Macmillian

Willetts, D., (1992) Modern Conservatism, London, Penguin

Grice, A., ‘Tory surge cuts Labour’s lead to five points’, The Independent, 15th March 2005.
